Iterative Monomorphisation

### 2025-02-13 iterative_monomorphisation.tex
it fits

\smallskip

@@

%Our findings are as follows:
%\begin{enumerate}
%\item Zipperposition benefits substantially from the E backend.
%
%\smallskip
%
%\item E with monomorphisation comes close second to the polymorphic prover Vampire.
%
%\smallskip
%
%\item For Leo-III~\cite{leo-iii} and Vampire~\cite{vamp}, we find that monomorphisation is indeed more effective than native polymorphism.
%\end{enumerate}

@@

\begin{examplex}Consider the following problem:
\begin{enumerate}
   \item \(\sym{p}\langle \ty{int}\rangle(0)\)
   \item \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){,}\; \sym{p}\langle\alpha\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
\end{enumerate}
%
The first iteration matches \(\alpha\) against \(\ty{int}\) for $\sym{p}$, generating the formula
%
\begin{enumerate}
   \item[3.] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){,}\; \sym{p}\langle\ty{int}\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
\end{enumerate}
%
The second iteration matches \(\alpha\) against \(\ty{list}(\ty{int})\), leading to the formula
%
\begin{enumerate}
   \item[4.] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){,}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
\end{enumerate}
%
Similarly the third iteration adds
%
\begin{enumerate}
   \item[5.] 
         \( \forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){,} \\ \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
\end{enumerate}

@@

The formula monomorphisation algorithm is given in Figure~\ref{mono_step}. It forms the core of the monomorphisation process by computing new type arguments tuples for a single formula. Type argument tuples are matched against each other to obtain a set of substitutions which is iterated over in the outermost loop. 
The separation of type arguments tuples into old and new maps is put to use to ensure that only combinations involving at least one new map are considered. This avoids re-computing some of the matchings processed in previous iterations. Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings. New type argument tuples are obtained by applying each substitution to each non-monomorphic type argument tuple such that at least one type variable in the tuple is instantiated by the substitution.

@@

   Depending on the shape and size of the input problem and the number of iterations performed, the doubly exponential growth may be problematic. Introducing bounds addresses this potential issue. The limit on the number of newly generated monomorphic type argument tuples is \(\min(\max(\MonoMult\cdot m, \MonoFloor), \MonoCap)\), where \(m\) is the total number of monomorphic type argument tuples. The components of this limit are
\begin{enumerate}
   \item \MonoCap, an absolute limit on the total number of new type argument tuples;
   \item \MonoMult, which is used to allow the total number of (monomorphic) type argument tuples to grow by a certain proportion of the current number \(m\) of monomorphic type argument tuples;
   \item \MonoFloor, which balances out \MonoMult, preventing \MonoMult from inhibiting new type argument tuple generation if \(m\) is too low.
\end{enumerate}

@@

Once all monomorphisation iterations have been completed, we are left with a set of the substitutions that have been used to generate new type arguments. The last phase uses these substitutions to instantiate the type variables in the input problem's non-monomrophic formuale. In the presence of bounds, the order in which the elements of the sets of substitutions of \(S\) are traversed affects the formulae resulting from the last phase. As we will discuss in Section~\ref{param_opti}, the best results were obtained when substitutions generated in the same iteration are grouped together.

@@

The Zipperposition implementation of the \emph{mono\_substs} function uses a map from type variables to substitutions instead of a set to filter the relevant substitutions from \(S\) more efficiently. The \textcolor{ourblueviolet}{max\_substs} bound exists for two main reasons:
\begin{enumerate}
   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic formula per initial polymorphic formula. Generating an excessive number of new formulae can flood the downstream prover. The final number of output formulae is ensured to be at most \(|\Phi|\cdot \text{\textcolor{ourblueviolet}{max\_substs}}\).
   \item The \emph{mono\_substs} function is the most explosive part of the algorithm. For a formula \(\varphi\), if \(S(\varphi)\) contains \(n\) substitutions that each instantiate exactly one of \(v\) different type variables, up to \(n^v\) monomorphising substititions may be generated. Recall that the total number of type argument tuples that are used to generate \(S(\varphi)\) can be doubly exponential in the number of loop iterations.
\end{enumerate}
%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]

@@

\begin{table}[ht]
\caption{Evaluation of Zipperposition without E vs.\ with E}

>>>>

@@

%Our findings are as follows:
%\begin{enumerate}
%\item Zipperposition benefits substantially from the E backend.
%
%\item E with monomorphisation comes close second to the polymorphic prover Vampire.
%
%\item For Leo-III~\cite{leo-iii} and Vampire~\cite{vamp}, we find that monomorphisation is indeed more effective than native polymorphism.
%\end{enumerate}

@@

\begin{examplex}\rm
Consider the following problem:
\begin{enumerate}
   \item \(\sym{p}\langle \ty{int}\rangle(0)\)
   \item \(\forall a: \alpha{,}\; \mathit{as}:\ty{list}(\alpha){,}\; \sym{p}\langle\alpha\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\alpha)\rangle(\mathit{as})\)
\end{enumerate}
%
The first iteration matches \(\alpha\) against \(\ty{int}\) for $\sym{p}$, generating the formula
%
\begin{enumerate}
   \item[3.] \(\forall a: \ty{int}{,}\; \mathit{as}:\ty{list}(\ty{int}){,}\; \sym{p}\langle\ty{int}\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{int})\rangle(\mathit{as})\)
\end{enumerate}
%
The second iteration matches \(\alpha\) against \(\ty{list}(\ty{int})\), leading to the formula
%
\begin{enumerate}
   \item[4.] \(\forall a: \ty{list}(\ty{int}){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{int})){,}\; \sym{p}\langle\ty{list}(\ty{int})\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{int}))\rangle(\mathit{as})\)
\end{enumerate}
%
Similarly the third iteration adds
%
\begin{enumerate}
   \item[5.] 
         \( \forall a: \ty{list}(\ty{list}(\ty{int})){,}\; \mathit{as}:\ty{list}(\ty{list}(\ty{list}(\ty{int}))){,} \\ \sym{p}\langle\ty{list}(\ty{list}(\ty{int}))\rangle(a) \rightarrow \sym{p}\langle \ty{list}(\ty{list}(\ty{list}(\ty{int})))\rangle(\mathit{as})\)
\end{enumerate}

@@

The formula monomorphisation algorithm is given in Figure~\ref{mono_step}. It forms the core of the monomorphisation process by computing new type arguments tuples for a single formula. Type argument tuples are matched against each other to obtain a set of substitutions which is iterated over in the outermost loop. 
The separation of type arguments tuples into old and new maps is put to use to ensure that only combinations involving at least one new map are considered. This avoids re-computing some of the matchings processed in previous iterations. %Not all such redundant computations can be avoided because the same substitution can be obtained from different matchings.
New type argument tuples are obtained by applying each substitution to each non-monomorphic type argument tuple such that at least one type variable in the tuple is instantiated by the substitution.

@@

   Depending on the shape and size of the input problem and the number of iterations performed, the doubly exponential growth may be problematic. Introducing bounds addresses this potential issue. The limit on the number of newly generated monomorphic type argument tuples is \(\min(\max(\MonoMult\cdot m, \MonoFloor), \MonoCap)\), where \(m\) is the total number of monomorphic type argument tuples. The components of this limit are
\begin{enumerate}
   \item \MonoCap, a %n absolute
   limit on the total number of new type argument tuples;
   \item \MonoMult, which is used to allow the total number of (monomorphic) type argument tuples to grow by a certain proportion of the current number \(m\) of monomorphic type argument tuples;
   \item \MonoFloor, which balances out \MonoMult, preventing \MonoMult from inhibiting new type argument tuple generation if \(m\) is too low.
\end{enumerate}

@@

Once all monomorphisation iterations have been completed, we are left with a set of the substitutions that have been used to generate new type arguments. The last phase uses this set to instantiate the type variables in the input problem's non-monomrophic formuale. In the presence of bounds, the order in which the elements of \(S\) are traversed affects the formulae resulting from the last phase. %As we will discuss in Section~\ref{param_opti}, the best results were obtained when substitutions generated in the same iteration are grouped together.

@@

The Zipperposition implementation of the \emph{mono\_substs} function uses a map from type variables to substitutions instead of a set to filter the relevant substitutions from \(S\) efficiently. The \textcolor{ourblueviolet}{max\_substs} bound exists for two main reasons:
\begin{enumerate}
   \item The iterative monomorphisation algorithm can generate up to \textcolor{ourblueviolet}{max\_substs} new monomorphic formula per initial polymorphic formula. Generating an excessive number of new formulae can flood the downstream prover. The final number of output formulae is ensured to be at most \(|\Phi|\cdot \text{\textcolor{ourblueviolet}{max\_substs}}\).
   \item The \emph{mono\_substs} function is the most explosive part of the algorithm. For a formula \(\varphi\), if \(S(\varphi)\) contains \(n\) substitutions that each instantiate exactly one of \(v\) %different
type variables, up to \(n^v\) monomorphising substititions may be generated. Recall that the total number of type argument tuples that are used to generate \(S(\varphi)\) can be doubly exponential in the number of loop iterations.
\end{enumerate}
%[This function can be thought of as exploring the tree of all possible monomorphising substitutions, \(S_{\text{res}}\) would represent the leaves that have already been visited and \(\sigma\) would represent the current node of the tree.]

@@

\begin{table}[t!]
\caption{Evaluation of Zipperposition without E vs.\ with E}
